### 2020-08-02 thesis/thesis.tex
[thesis] Many small improvements

%%%%% DANE DO STRONY TYTUŁOWEJ
\polishtitle{Efekty koalgebraiczne oraz ich kohandlery \fmlinebreak{} w językach programowania}
\englishtitle{Coalgebraic effects and their cohandlers \fmlinebreak{} in programming languages}
\polishabstract{
    \ldots
}
\englishabstract{
    \ldots
}
\author{Mateusz Urbańczyk}
\advisor{dr Maciej Piróg}
\date{1 września 2020}                     % Data zlozenia pracy
% Dane do oswiadczenia o autorskim wykonaniu
% \transcriptnum {291480}                     % Numer indeksu
% \advisorgen    {dr. Macieja Piróga} % Nazwisko promotora w dopelniaczu
%%%%%

@@

% ...
%%%%%

@@

% TODO This requires a rework

In this thesis we present experimental programming language Freak, which is an
implementation of Continuation Passing Style for Effect Handlers paper~\cite{handlers-cps},
with the addition of a few basic constructs. We start by presenting the related work,
then discuss syntax and operational semantics. Basic usage guide for playing with the
language is provided, as well as implementation details and examples. We conclude by
stating what are the possible augmentations, that are intended to be made in the future.

@@

What are the issues with normal algebraic effects?

@@

Brief and concise description of the problem

@@

@@

Under these circumstances one may pose a question, why do we still bother with
development of languages theory, since so much has been done already. Is there
anything that drives us towards further research? Indeed, one active branch
revolves around equational theory to asses equality of two programs, which we
know that in general setting is undecidable. Proof methods may include extensional,
contextual or logical equality. However, there is no doubt that these formal
ways of reasoning about programs, while being crucial for assessing correctness,
do not bring direct benefits for practical, everyday use cases.

% Not sure I like this part. Isn't it too broad to say handle complexity?
Other branch of languages theory, that we shall investigate more in this thesis,
is about handling complexity of programs. Various methods of static analysis has
been developed, most notably, type systems.

What is the source of complexity? Thanks to strong and static type systems along
with their implementations, we have solid tools to work efficiently on functions
that are pure. That being said, we claim that the sole complexity of programs
comes from side effects, which we cannot avoid in writing anything useful.

We need to have a good way for handling computational effects, which so far has
been modeled through monads~\cite{moggi}.

However, they were found to be, to say the least, cumbersome to work with when
complexity and number of different effects increases. It is perhaps not a coincidence
that many functional programming languages do not have them, because of the
non-composable nature of them.

Having functions $ f: a \rightarrow m b $ and $ g : b \rightarrow m' c $, it
should not be an issue to compose them together.

Recall that the main idea of monads lies on the fact that we provide a function
to lift value into a functor and associative kleisli composition operator to combine
computations together. It turns out, that it becomes complex and unpleasent to
combine two functors together to form a new monad. For this purpose, monad
transformers arose in Haskell.

Most of the common languages avoid this by not expressing computational effects
in the type system, and instead one may think about functions as being implicitly
embed in a Kleisli Category over a functor T, where T is a hidden signature over
all possible side effects that occur in our program.

Not only we would like to bring back effects to our type system, but also do it
in a way that is composable. This is where algebraic effects comes to the rescue.

From theoretical point of view, we need to develop equa about our effects
to assert correctness of our langauge as well as to have the right hammer
to reason about our programs.

That is the point where we would like to introduce to unfamiliar read a notion
of algebraic effects.

@@

What are algebraic effects?

% Algebraic effects as Interfaces for side effects

@@

@@

Algebraic effects can be describe via operational means, however,
for the purpose of presenting the duality between algebra and coalgebra,
we allow ourselves to wander a bit deeper into category theory and describe
effects from denotational point of view.

@@

    Collection of $\mathVar{\Sigma-terms}$ is a free algebra with a generator $X$
    for a functor $ \mu H_{\Sigma} $ that maps objects into trees over a given
    signature $ \Sigma $ and morphisms into folds over trees.

@@

    An \textit{algebraic theory} $T = (\Sigma_{T}, \mathcal{E}_{T})$, is given by
    a signature $\Sigma_{T}$ and a collection $\mathcal{E}_{T}$ of
    $\mathVar{\Sigma_{T}-equations}$. For clarity we will usually omit T subscript

@@

    \begin{center}
    \begin{tikzcd}[row sep=huge, column sep=large]
        {|M|}^k \arrow[rr, '\phi', bend right] \arrow[rr, '\rrbracket{} X|l', bend left] &  & {|M|}
    \end{tikzcd}
    \end{center}
    % TODO: this diagram doesn't work

@@

    % TODO:
    \begin{center}
    \begin{tikzcd}[row sep=huge, column sep=large]
    % F A^{\#} \arrow[d, "F \overline{f}"] \arrow[r, "\varphi_{A}"] & A^{\#} \arrow[d, "\overline{f}"] & A \arrow[l, "\eta_{A}"'] \arrow[ld, "f"] \\
    % F B \arrow[r, "\beta"']                                       & B                                &
    \end{tikzcd}
    \end{center}

@@

    % TODO

@@

    Comodel in $\mathcal{C}$ is just a Model in $ \mathcal{C}^{op} $.

    Models --- Worlds

@@

    Coinduction, and coinductive structures, provide us a way of observation
    of the behaviour, as a contractry to construction in inductive reasoning.
    Simple example of the duality can be expressed through induction over finite
    list and coinductive observation of infinite streams.

@@

@@

    % TODO: On it's own this snippet a bit dry and of course requires additional examples,
    % for instance the classic one with operating on closed file

@@

@@

One-shot continuations

Detecting whether resumption is called only once is undecidable.

@@

@@

@@

@@

\chapter{Effectful and Coeffectful Programming}
\section{Examples}\label{sec:examples}

@@

    The first example will be based on modelling (nondeterministic) choice
    in the program. We will make two decisions, which will affect the computation
    result:

@@

@@

@@

@@

@@

>>>>

\polishtitle{Efekty koalgebraiczne oraz ich kohandlery \fmlinebreak{} w językach programowania}
\englishtitle{Coalgebraic effects and their cohandlers \fmlinebreak{} in programming languages}
\polishabstract{
    \ldots
}
\englishabstract{
    \ldots
}
\author{Mateusz Urbańczyk}
\advisor{dr Maciej Piróg}
\date{1 września 2020}                     % Data zlozenia pracy
% Dane do oswiadczenia o autorskim wykonaniu
% \transcriptnum {291480}                     % Numer indeksu
% \advisorgen    {dr. Macieja Piróga} % Nazwisko promotora w dopelniaczu
%%%%%

@@

@@

TODO This requires a rework

In this thesis we present experimental programming language Freak, which is an
implementation of Continuation Passing Style for Effect Handlers paper~\cite{handlers-cps},
with the addition of a few basic constructs.

@@

TODO add citations

Algebraic effects, while not being a new concept, in the recent years
have received a lot of attention, both from theoretical and practical side.

TODO refactor the following

They give developer necessary tools for declarative programming, writing
programs in a way to strive from exposing implementation details, while still
preserving a lot of flexibility and having strong theoretical background and
ways of formal reasoning.

In fact, that very flexibility can be troublesome. Allowing developer for too
much freedome may lead to undesired behaviour and incorrect programs, for that
very reason, we develop type systems, to detect errors before they occur.

@@

TODO Think about merging Problem Analysis with Statement \\

Multi-shot resumptions in effect handlers are a powerful tool to describe
fairly complex logic in a concise way. That being said, they give rise to
issues that would not occur in standard control flow, especially when composing
various effects together.

We will explain this in greater details and think about possible ways to address
the issue of flexibility of effects in the next sections.

TODO actually point out which sections you talk about

@@

TODO Guide reader through thesis, probably later than sooner \\
TODO Clearly divide categorical from operational approach for less algebraically fluent readers

TODO it's outdated
We start by presenting the related work,
then discuss syntax and operational semantics. Basic usage guide for playing with the
language is provided, as well as implementation details and examples. We conclude by
stating what are the possible augmentations, that are intended to be made in the future.

@@

Under these circumstances one may pose a question, why do we still bother with
development of languages theory, since so much has been done already. Is there
anything that drives us towards further research? Indeed, one active branch
revolves around equational theory to asses equality of two programs, which we
know that in general setting is undecidable. Proof methods may include extensional,
contextual or logical equality. However, there is no doubt that these formal
ways of reasoning about programs, while being crucial for assessing correctness,
do not bring direct benefits for everyday use cases, as they are rarely accessible
by a common developer.

Other branch of languages theory, that we shall investigate more in this thesis,
is about taming complexity of programs. Various methods of static analysis has
been developed for various use cases, most notably, type systems.

What is the source of complexity? Thanks to strong and static type systems along
with their implementations, we have solid tools to work efficiently on functions
that are pure. That being said, we claim that the core complexity of programs
comes from side effects, which we cannot avoid in writing anything useful.

We need to have a good way for handling computational effects. One of the ways
to model them, can be done through monads~\cite{moggi}. However, they were
found to be, to say the least, cumbersome to work with when the number of
different effects increases. It is perhaps not a coincidence that many
functional programming languages do not have them, because of the non-composable
nature of them, or at least not composable in their implementations.

Let's put the following functions

    $ f: a \rightarrow b $, $ g : b \rightarrow c $ \\
    $ f': a \rightarrow m \; b $, $ g' : b \rightarrow m \; c $ \\
    $ f'': a \rightarrow m' \; m \; b $, $ g'' : b \rightarrow m' \; m \; c $ \\

\end{center}

where $m$, $m'$ are monads. Functions $f$, $g$ can be composed using standard
composition, for $f'$, $g'$ we can use Kleisli composition operator from monad $m$.
What in case of adding one more effect?

It turns out, that it becomes complex and unpleasent to combine two functors
together to form a new monad, for which purpose, monad transformers arose in Haskell.
Most of the common languages avoid this by not expressing computational effects
in the type system, and instead one may think about functions as being implicitly
embed in a Kleisli Category over a functor T, where T is a hidden signature over
all possible side effects that occur in our program.

TODO Think about describing what is Kleisli Category

Not only we would like to bring back effects to our type system, but also do it
in a way that is composable. This is where algebraic effects comes to the rescue.
From theoretical point of view, we need to develop equational theory about our
effects to assert correctness of our langauge as well as to have the right hammer
to reason about our programs. That is the point where we would like to introduce
to unfamiliar reader a notion of algebraic effects.

@@

@@

Patterns like these are well known in programming, for which other alternatives
arose. One can mention interfaces from object-oriented programming, which are also
a way to describe the API of a certain component. In order to abstract from
implementation details, we pass an object around which represents a certain
interface. While this is certainly better than having no abstractions at all,
we need to pass this interface into every function that is going to use it.
This practice is called dependency injection, and while it sounds like it solves
some of the issues, we end up in functions that need to carry representants
of the interfaces, and pass them in every subcall that they make. It also can't
be ommited that it's not really a substitute, as it's solving only one particular
issue that algebraic effects address.

TODO add link to paper with plenty of examples

- Exceptions
- Async/await
- Flipping coin
- Backtracking?
- Operating on file

@@

Algebraic effects can be described via operational means, however,
for the purpose of presenting the duality between algebra and coalgebra,
we allow ourselves to wander a bit deeper into category theory and describe
effects from denotational point of view.

@@

    \begin{definition}
    Collection of $\mathVar{\Sigma-terms}$ is a free algebra with a generator $X$
    for a functor $ \mu H_{\Sigma} $ that maps objects into trees over a given
    signature $ \Sigma $ and morphisms into folds over trees.

@@

    An \textit{algebraic theory} $T = (\Sigma_{T}, \mathcal{E}_{T})$, is given
    by a signature $\Sigma_{T}$ and a collection $\mathcal{E}_{T}$ of
    $\mathVar{\Sigma_{T}-equations}$. For clarity, we will usually omit T subscript.

@@

    \begin{center}
    \begin{tikzcd}[row sep=huge, column sep=large]
        {|M|}^k \arrow[rr, '\phi', bend right] \arrow[rr, '\rrbracket{} X|l', bend left] &  & {|M|}
    \end{tikzcd}
    \end{center}

    TODO this diagram doesn't work

@@

    TODO we can also say it's an initial algebra over initial object.

    \begin{center}
    \begin{tikzcd}[row sep=huge, column sep=large]
    % F A^{\#} \arrow[d, "F \overline{f}"] \arrow[r, "\varphi_{A}"] & A^{\#} \arrow[d, "\overline{f}"] & A \arrow[l, "\eta_{A}"'] \arrow[ld, "f"] \\
    % F B \arrow[r, "\beta"']                                       & B                                &
    \end{tikzcd}
    \end{center}

@@

    TODO finish definition

@@

    \begin{definition}
        Comodel in $\mathcal{C}$ is just a Model in $ \mathcal{C}^{op} $.
    \end{definition}

    We could end by stating this definition, and everything else would follow
    directly from duality. However, expanding definitions is going to give us
    better intuition, as well as give more solid ground for implementation.

@@

    Induction and Coinduction

    Induction is a way of constructing new structures. Recursion is a way of
    folding inductively defined structure in an terminating way. Recursive functions
    should shrink the argument in each call, meaning that it eventually ends up
    terminating in a base case.

    Coinduction is literally a dual notion to induction. We \textit{observe} possibly
    infinite structures, by doing deconstruction. Corecursion, however, is a way
    of productively defining new, possibly enlarged structures. Due to infinity,
    the evaluation should be lazy, whereas in induction it may be eager.

    Here is a table that summarizes difference between these two methods of reasoning:

    TODO rewrite into latex-table \\

    | feature             | induction            | coinduction          |

    |---------------------|----------------------|----------------------|

    | basic activity      | construction         | deconstruction       |

    | derived activity    | deconstruction       | construction         |

    | functions shape     | inductive domain     | coinductive codomain |

    | (co) recursive calls| shrinks the argument | grows the result     |

    | functions feature   | terminating          | productive           |

    | evaluation          | possibly eager       | necessarily lazy     |

    Reader that focuses on pragmatism may pose a question, why do we even want
    to reason about infinite structures? They never appear in practice! In fact,
    it's very common to operate on never-ending transition systems or streams of
    data without, wher efinitary means of reasoning are of no use, as we can’t
    expect an end to stream!

    Simple example of the duality can be expressed through induction over finite
    list and coinductive observation of infinite streams.

@@

    TODO Write a coinductive proof for divergent terms using coinductive proof
    principle.

\subsection{Coinduction Coalgebra something section}

    TODO Rethink this section, back by examples

@@

    TODO On it's own this snippet a bit dry and of course requires additional examples,
    for instance the classic one with operating on closed file and/or on
    streams of data

@@

    In fact, libraries for algebraic effects also arose in mainstream languages,
    such as C or Python. In Python, effects are implemented through generators,
    using built-in feature of sending value when doing \textit{yield} operation.
    Resumptions in handlers that are sending value, are one-shot and tail-recursive,
    therefore we do not need to handle coalgebraic part, at the cost of flexibility.

    TODO add reference to C and Python lib

@@

TODO smth smt initialization before finalization.

It can be seen, that issues lies with the fact, that interacting with coalgebras
may lead to change of their internal state system. Examples include change
of state in NFA, taking next element in infinite stream or a closing file.

One of the ways to approach this problem, caused by extensive generality of
effects, is extension of the calculus with Coalgebraic effects, also named
coeffects. Difference is that resumptions in coeffects handlers are one-shot
only, which means that continuation can be invoked at most once.

\begin{lemma}
    Problem of detection whether resumption is called only once is undecidable.
\end{lemma}

\begin{proof}
    Follows directly from Rice's Theorem, as checking whether function was called
    is a nontrivial semantic property of a program.
\end{proof}

@@

One approach is to dynamically during execution check against contract that
continuation may only be called once. In this setting we are sure that our
program will not accidentaly go into wrong state, however by definition we lack
static analysis to prevent from errors before they occur.

@@

TODO Explain it can get complex and unfamiliar for users.

@@

TODO Describe what is data-flow analysis \\

TODO Give example for constant folding \\

TODO Propose analysis on imaginary CFG Not sure if it makes sense?

@@

\chapter{(Co) Effectful Programming}  % TODO w/o space?
\section{Examples}\label{sec:examples}

@@

    The first example will be based on modelling (nondeterministic) choice
    in the program. We will make two decisions, which will affect the computation

    result:

@@

    TODO subset of this section's content is explained already

@@

    TODO Extend language with basic constructs for usability

@@

    TODO Rewrite here operational semantics from Links paper

@@

    TODO Drop of continuations is still an issue

@@

    TODO Simple implementation for cohandlers
